Mixture Models for Single-Cell Assays with Applications to Vaccine Studies

### 2013-06-25 MIMOSA.tex
update paper with 2nd round of revisions

Cell counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences. \emph{Responders} are subjects whose T-cells respond to stimulation, and the response is vaccine-specific if it exists at the post-vaccine time point but not the pre-vaccine time point.
%Response is vaccine--specific when a subject responds at a post--vaccine time point but not at a pre--vaccine time point. 
%Response in placebo recipients is further used to assess vaccine-specificity. 
These comparisons are typically performed within a time point in order to avoid confounding vaccine effects time and because baseline samples are not always available. 
%In many immunological studies, the number of positive cells in the functionally distinct subpopulations is very small (on the order of 10s of cells), with correspondingly small effect sizes, thus real biological differences are difficult to detect.

@@

\section{Differential expression with one biomarker}
\label{s:DEone}
Datasets like those presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many combinations. By univariate, we mean that we have only one positive cell subset. This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ or B$+$ (\textit{vs.} A$-$/B$-$). Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$). In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the unstimulated and stimulated samples (Table \ref{tab:twobytwo}).

@@

For a given subject and stimulation, we consider a biomarker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the proportion of positive cells in the unstimulated sample. Subjects that show differential expression will be called responders for that biomarker. In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model.
\subsection{Beta-binomial model}
\label{s:DE}
For a given subject $i$, the positive cell counts for the stimulated and unstimulated samples are jointly modeled as $(n_{si}|p_{si}) \sim \mathrm{Bin}(N^{(s)}_{i},p^{(s)}_{i})\quad \text{and}\quad (n^{(u)}_{i}|p^{(u)}_{i}) \sim \mathrm{Bin}(N^{(u)}_{i},p^{(u)}_{i})\label{eq:bino_likelihood}$ where $p^{(s)}_{i}$, $p^{(u)}_{i}$ are the unknown proportions for the stimulated and unstimulated paired samples, respectively. In order to detect responding subjects, we consider two competing models,
${\cal M}_0: p^{(u)}_{i}=p^{(s)}_{i}$ and ${\cal M}_1: p^{(u)}_{i}\ne p^{(s)}_{i}. \label{eq:models}$
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and unstimulated samples, and the proportions are equal (yet the cell counts can differ). Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder. In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p^{(s)}>p^{(u)}$. This alternative parametrization is described in Appendix A of the Supplementary Material, and we refer to it as the one-sided model.

@@

\section{Results}
\label{s:results}
We evaluated and compared the performance of MIMOSA against Fisher's exact test, a likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by compared the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for the ICS data.  A false discovery is defined as a day 0 sample or a sample from a placebo recipient (expected non-responders) that is identified as a responder by an algorithm. For MIMOSA, the FDR was computed using the approach of~\citet{Newton2004} whereas for all other methods FDR control was done using the approach of~\citet{Benjamini1995}.

\subsection{ICS}
\label{s:ics}
In the ICS data we considered observations from vaccinees at day 0 or from placebos at day 0 or 182 as \emph{true negatives} in that we do not expect to observe a response in those samples. Observations from vaccinees at day 182 were considered \emph{true positives} in that a response is expected assuming a 100\% response rate. This assumption was required, since we do not have independent, external measurement of the true response. Under these assumptions, the MIMOSA model has higher sensitivity and specificity than method compared here for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positives being detected.
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F). Here we present the results based on IL2 and IFN-$\gamma$ alone as well as the subset expressing IL2 or IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}. These results are consistent for other cytokines and cytokine combinations, including higher order combinations  (Supplementary Figure 1).  We also evaluated, through simulation, what would happen to the ROC curves when our assumption of 100\% response rate was false (\textit{i.e.,} when not all day 182 vaccinees are responders, Supplementary Figure 2), and found that we are potentially underestimating the sensitivity of all methods considered here due to non-responders at day 182. Nonetheless, the ranking of the methods remained consistent and the conclusions from the ROC analysis still hold.

@@

\subsection{Single-cell gene expression}
We applied the MIMOSA model to the Fluidigm single-cell gene expression data set. We used the two-sided MIMOSA model because genes could be regulated upward or downward upon stimulation. In order to detect stimulation specific changes of expression, we fit our model separately to each gene within each stimulation. The results show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A). These patterns evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A) closely match the empirical estimates of the differences in proportions (Figure~\ref{fig:fluidigm} B). As a comparison, we also fit the model combining information across all stimulations (Figure~\ref{fig:fluidigm} C). Although this leads to a sensible clustering by stimulation, the observed pattern does not agree as well with the empirical differences. A similar analysis applying a two-sided Fisher's exact test to each sample and gene, and clustering the signed FDR adjust q-values (Figure~\ref{fig:fluidigm} D) does not reveal stimulation-specific patterns, with the exception of CMV pp65 nlv5 stimulation. At an FDR of 10\%, Fisher's exact test identified 13 significant genes, while MIMOSA identified 46 significant genes. The 13 genes identified by Fisher's method were a subset of those identified by MIMOSA. These results suggest that fitting a stimulation--specific model is the right approach since information is shared across subjects that are expected to exhibit similar behavior.

@@

\subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
As a proof-of-concept, we applied our multivariate MIMOSA model to two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example $K=2$ and we have four possible subsets.
In Figure~\ref{fig:polyfunctionality} we show heatmaps of the $\log_2$ counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples, as well as for the $log_2$ of the sum of the counts of the 3 positive subsets. The number of CCR7-GZML+ cells increases upon stimulation, while the number of CCR7+GZMK- cells decreases. The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all polyfunctional cell populations (\textit{i.e.}, cells expressing CCF7$+$ or GZMK$+$), would not be appropriate in this case, since changes in the counts are bidirectional and their sum is near zero. The multivariate MIMOSA model tests for any difference across all polyfunctional cell subsets and detects significant differences in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, dark labels), whereas the standard approach detects no differences (Figure~\ref{fig:polyfunctionality} B). Testing all combinations simultaneously is also an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.

@@

All of the results presented here were obtained with an R and C++ implementation the MIMOSA models, freely available at (\url{http://www.github.org/RGLab/MIMOSA}).

@@

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.75\columnwidth]{FIG1.eps}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2. E-F) ENV-1-PTEG and expressing IFN-$\gamma$ or IL2. ROC and FDR plots of other cytokine combinations can be found in Supplementary Figure 1.This figure appears in color in the electronic version of this article.}
\label{fig:HVTN065}
\end{figure}

\begin{figure}
\centering
\includegraphics{TIKZFig2.eps}
%\begin{tikzpicture} [auto,node distance=0cm]
%\node at (0,0) (A) {
%\begin{tikzpicture}
%\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot1.pdf}};
%\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
%\node at (0,1) [font=\small\sffamily] {A} ;
%\end{scope}
%\end{tikzpicture}
%};
%\node [right=of A] (B) {
%\begin{tikzpicture}
%\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot2.pdf}};
%\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
%\node at (-0.1,1.1) [font=\small\sffamily] {B} ;
%\end{scope}
%\end{tikzpicture}
%};
%\node [below=of A] (C) {
%\begin{tikzpicture}
%\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.45\columnwidth]{Figures/FluidigmFisher.pdf}};
%\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
%\node at (-0.05,1) [font=\small\sffamily] {C} ;
%\end{scope}
%\end{tikzpicture}
%};
%\end{tikzpicture}
\caption{Signed posterior probability and difference in empirical  in proportions of single-cells expressing each gene on a 96x96 Fluidigm array. The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control) for a model fit to each stimulation and gene separately. Columns and rows are clustered based on these signed posterior probabilities. B) The empirical differences in proportion of cells expressing a gene in the stimulated vs. control samples. Rows and columns are ordered as in A) for comparison. The traces show the deviations of each cell from zero. Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10). C) Signed posterior probabilities for a model fit per gene but across all stimulations simultaneously. D) Clustering of the signed q-values from Fisher's exact test. All genes  were significant at, or below, the 10\% FDR level. This figure appears in color in the electronic version of this article.}
\label{fig:fluidigm}
\end{figure}

>>>>

Cell counts are compared between antigen stimulated and unstimulated samples from a subject to identify significant differences. \emph{Responders} are subjects whose T-cells respond to stimulation, and the response is vaccine-specific if it exists at the post-vaccine time point but not the pre-vaccine time point.
%Response is vaccine--specific when a subject responds at a post--vaccine time point but not at a pre--vaccine time point. 
%Response in placebo recipients is further used to assess vaccine-specificity. 
These comparisons are typically performed within a time point in order to avoid confounding vaccine effects with time of visit and because baseline samples are not always available. 
%In many immunological studies, the number of positive cells in the functionally distinct subpopulations is very small (on the order of 10s of cells), with correspondingly small effect sizes, thus real biological differences are difficult to detect.

@@

\section{Differential expression with one biomarker}
\label{s:DEone}

Datasets like those presented here are usually analyzed in a univariate fashion to avoid being underpowered due to the large number of combinations and the potential for very small cell counts in many combinations. By univariate, we mean that we have only one positive cell subset. This cell subset can be defined by considering the expression of one biomarker alone (marginalizing over all other measured biomarkers) such as A$+$ (\textit{vs.} A$-$), or considering a specific positive biomarker combination (and marginalizing over everything else) such as A$+$ OR B$+$ (\textit{vs.} A$-$/B$-$). Without loss of generality, we treat the univariate case as a one biomarker case (\textit{i.e.}, $K=1$). In this case, for a given subject, the data can be summarized in a contingency table of $+/-$ cell counts across the unstimulated and stimulated samples (Table \ref{tab:twobytwo}).

@@

For a given subject and stimulation, we consider a biomarker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the proportion of positive cells in the unstimulated sample. Subjects that show differential expression will be called responders for that biomarker. In this section, we are concerned with identifying differential expression one biomarker at a time, using a beta-binomial mixture model.
\subsection{Beta-binomial model}
\label{s:DE}

For a given subject $i$, the positive cell counts for the stimulated and unstimulated samples are jointly modeled as $(n^{(s)}_{i}|p^{(s)}_{i}) \sim \mathrm{Bin}(N^{(s)}_{i},p^{(s)}_{i})\quad \text{and}\quad (n^{(u)}_{i}|p^{(u)}_{i}) \sim \mathrm{Bin}(N^{(u)}_{i},p^{(u)}_{i})\label{eq:bino_likelihood}$ where $p^{(s)}_{i}$, $p^{(u)}_{i}$ are the unknown proportions for the stimulated and unstimulated paired samples, respectively. In order to detect responding subjects, we consider two competing models,
${\cal M}_0: p^{(u)}_{i}=p^{(s)}_{i}$ and ${\cal M}_1: p^{(u)}_{i}\ne p^{(s)}_{i}. \label{eq:models}$
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and unstimulated samples, and the proportions are equal (yet the cell counts can differ). Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the subject $i$ is a responder. In some studies, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p^{(s)}>p^{(u)}$. This alternative parametrization is described in Appendix A of the Supplementary Material, and we refer to it as the one-sided model.

@@

\section{Results}
\label{s:results}
We evaluated and compared the performance of MIMOSA against Fisher's exact test, a likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by compared the \textit{observed FDR} (false discovery rate) against the \textit{nominal FDR} (expected false discovery rate) for the ICS data.  A false discovery is defined as a day 0 sample or a sample from a placebo recipient (expected non-responders) that is identified as a responder by an algorithm. For MIMOSA, the FDR was computed using the approach of~\citet{Newton2004} whereas for all other methods FDR control was done using the approach of~\citet{Benjamini1995}. The posterior mean of the $z_i$ was used for ranking the results in the FDR calculations.

\subsection{ICS}
\label{s:ics}
In applying the MIMOSA model to the ICS data set, we fit a separate model for each cytokine subset (and each stimulation if we had analyzed more than one).  We do not compare pre-vaccination and post-vaccination time points against each other directly. Rather, we have paired samples for each time points with the corresponding negative controls. Observations within each time point are contrasted against the corresponding negative controls, and we fit the model to observations from all time points simultaneously. Consequently all time points that are fit simultaneously will contribute to the priors for $p^{(u)}$. A subsequent evaluation of pre- and post-vaccination time points would involve comparing the positivity calls for paired samples at the two time points. However, data are not typically analyzed this way, as this could lead to confounding of date with vaccine effect (i.e. batch effects). Furthermore, vaccine trials do not always collect baseline samples.

Within this framework, we considered observations from vaccinees at day 0 or from placebos at day 0 or 182 as \emph{true negatives} in that we do not expect to observe a response in those samples. Observations from vaccinees at day 182 were considered \emph{true positives} in that a response is expected assuming a 100\% response rate. This assumption was required, since we do not have independent, external measurement of the true response. Under these assumptions, the MIMOSA model has higher sensitivity and specificity than method compared here for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065}, panels A,C,E.
At an FDR between 10-20\%, MIMOSA would lead to about 20\% more true positives being detected.
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065}, panels B,D,F). Here we present the results based on IL2 and IFN-$\gamma$ alone as well as the subset expressing IL2 OR IFN-$\gamma$ that were used in the original study~\citep{Goepfert:2011ci}. These results are consistent for other cytokines and cytokine combinations, including higher order combinations  (Supplementary Figure 1).  We also evaluated, through simulation, what would happen to the ROC curves when our assumption of 100\% response rate was false (\textit{i.e.,} when not all day 182 vaccinees are responders, Supplementary Figure 2), and found that we are potentially underestimating the sensitivity of all methods considered here due to non-responders at day 182. Nonetheless, the ranking of the methods remained consistent and the conclusions from the ROC analysis still hold.

@@

\subsection{Single-cell gene expression}
We applied the MIMOSA model to the Fluidigm single-cell gene expression data set. We used the two-sided MIMOSA model because genes could be regulated upward or downward upon stimulation. In order to detect stimulation specific changes of expression, we fit our model separately to each gene within each stimulation. The results show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene while preserving inter-subject variability (Figure~\ref{fig:fluidigm} A). These patterns evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A) closely match the empirical estimates of the differences in proportions (Figure~\ref{fig:fluidigm} B). As a comparison, we also fit the model combining information across all stimulations (Figure~\ref{fig:fluidigm} C). Although this leads to a sensible clustering by stimulation, the observed pattern does not agree as well with the empirical differences. A similar analysis applying a two-sided Fisher's exact test to each sample and gene, and clustering the signed FDR adjust q-values (Figure~\ref{fig:fluidigm} D) does not reveal stimulation-specific patterns, with the exception of CMV pp65 nlv5 stimulation.  We note that the differences in the clustering are not surprising given that the MIMOSA model is sharing information across subjects within a stimulation. At an FDR of 10\%, Fisher's exact test identified 13 significant genes, while MIMOSA identified 46 significant genes. The 13 genes identified by Fisher's method were a subset of those identified by MIMOSA. These results suggest that fitting a stimulation--specific model is the right approach since information is shared across subjects that are expected to exhibit similar behavior.

@@

\subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
As a proof-of-concept, we applied our multivariate MIMOSA model to two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example $K=2$ and we have four possible subsets.
In Figure~\ref{fig:polyfunctionality} we show heatmaps of the $\log_2$ counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples, as well as for the $log_2$ of the sum of the counts of the 3 positive subsets. The number of CCR7-GZML+ cells increases upon stimulation, while the number of CCR7+GZMK- cells decreases. The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all polyfunctional cell populations (\textit{i.e.}, cells expressing CCF7$+$ OR GZMK$+$), would not be appropriate in this case, since changes in the counts are bidirectional and their sum is near zero. The multivariate MIMOSA model tests for any difference across all polyfunctional cell subsets and detects significant differences in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, dark labels), whereas the standard approach detects no differences (Figure~\ref{fig:polyfunctionality} B). Testing all combinations simultaneously is also an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.

@@

All of the results presented here were obtained with an R and C++ implementation the MIMOSA models, freely available at (\url{http://www.github.org/RGLab/MIMOSA}). The code to replicate the analyisis is available at (\url{http://www.github.org/RGLab/MIMOSA_manuscript/MIMOSA/mimosafigures}).

@@

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.75\columnwidth]{FIG1.eps}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2. E-F) ENV-1-PTEG and expressing IFN-$\gamma$ OR IL2. ROC and FDR plots of other cytokine combinations can be found in Supplementary Figure 1.This figure appears in color in the electronic version of this article.}
\label{fig:HVTN065}
\end{figure}

\begin{figure}
\centering
\includegraphics{TIKZFig2.eps}
%\begin{tikzpicture} [auto,node distance=0cm]
%\node at (0,0) (A) {
%\begin{tikzpicture}
%\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot1.pdf}};
%\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
%\node at (0,1) [font=\small\sffamily] {A} ;
%\end{scope}
%\end{tikzpicture}
%};
%\node [right=of A] (B) {
%\begin{tikzpicture}
%\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.45\columnwidth]{Figures/Fluidigm_plot2.pdf}};
%\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
%\node at (-0.1,1.1) [font=\small\sffamily] {B} ;
%\end{scope}
%\end{tikzpicture}
%};
%\node [below=of A] (C) {
%\begin{tikzpicture}
%\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.45\columnwidth]{Figures/FluidigmFisher.pdf}};
%\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
%\node at (-0.05,1) [font=\small\sffamily] {C} ;
%\end{scope}
%\end{tikzpicture}
%};
%\end{tikzpicture}
\caption{Signed posterior probability and difference in empirical  in proportions of single-cells expressing each gene on a 96x96 Fluidigm array. The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control) for a model fit to each stimulation and gene separately. Columns and rows are clustered based on these signed posterior probabilities. B) The empirical differences in proportion of cells expressing a gene in the stimulated vs. control samples. Rows and columns are ordered as in A) for comparison. The traces show the deviations of each cell from zero. Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10). C) Signed posterior probabilities for a model fit per gene but across all stimulations simultaneously. D) Clustering of the signed q-values from Fisher's exact test. The differences in clustering are not surprising given that the MIMOSA model shares information within stimulation. All genes  were significant at, or below, the 10\% FDR level. This figure appears in color in the electronic version of this article.}
\label{fig:fluidigm}
\end{figure}
